---
title: "COVID-19-Analysis"
author: "Jessica W"
date: "2024-04-08"
output: html_document
---

# Introduction

The COVID-19 pandemic made a significant impact on global health and societies.
Understanding how the virus and how it impacted various areas and their
populations is very important in shaping the future of public health policy
decisions and how vulnerable certain areas can be. In this analysis, we will
examine the spread of the data throughout 2020-2023 in the United States using
publicly available data from John's Hopkins.

Our objective in this analysis is to explore the trend of cases and deaths from
COVID-19 through the years provided by the data and what the propagation of
cases looked like throughout the United States. Then specifically, we will take
a closer look at the spread of cases is in my own state of Wisconsin across its
cities.

We'll begin by cleaning and pre-processing the data, followed by exploratory
data analysis and visualization to understand the distribution and trends of 
COVID-19cases and deaths throughout both the US as a whole as well as a closer
look at one state.

Through this analysis, we hope to contribute to the ongoing efforts to combat
the COVID-19 pandemic and mitigate its impact on communities around the world.


# Setup and Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries, warning= FALSE, message = FALSE}

# Import necessary libraries
library(tidyverse)
library(tidyr)
library(lubridate)
library(RColorBrewer)
library(leaflet)
library(geojsonio)
```


# Data Preparation


```{r data import, message = FALSE}

# Initial constant url
url_in <- 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/'

# Files needed
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

# Concatenate strings
urls <- str_c(url_in, file_names)

# Read in CSVs
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

### Might not Need ###
lookup_table <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")
```


```{r states population}
# View of the lookup table that has been filtered to only the 50 States
head(lookup_table %>% 
       filter(iso3 == 'USA'), 60)

# Filter out any unwanted entries to only get entries that include population
# for the whole state
us_states_pop <- lookup_table %>%
  filter(
    iso3 == 'USA' &
    is.na(Admin2) &
    !is.na(Population) &
    Province_State >= "Alabama" &
    Province_State <= "Wyoming"
  )

# Verify the data for later use
us_states_pop
```



```{r cases/deaths tables}

# Look at if there are any areas/regions outside of the US like Puerto Rico etc
us_cases %>% 
  count(iso3)

# Filter to only the States for this analysis
us_cases <- us_cases %>% 
  filter(iso3 == 'USA')

us_deaths <- us_deaths %>% 
  filter(iso3 == 'USA')

# View the data
head(us_cases)
head(us_deaths)
```


```{r cases/deaths clean up}

# Filter out unnecessary columns and pivot the dates to a Date column and create
# new column for cases on that day
us_cases <- us_cases %>%
  select(-c('UID', 'iso2', 'iso3', 'code3', 'FIPS', 'Country_Region', 'Combined_Key')) %>%
  pivot_longer(cols = -c('Admin2','Province_State', 'Lat', 'Long_'),
               names_to = 'Date',
               values_to = 'cases') %>%
  mutate(Date = mdy(Date))


head(us_cases)

# Same process for the US deaths
us_deaths <- us_deaths %>%
  select(-c('UID', 'iso2', 'iso3', 'code3', 'FIPS', 'Country_Region', 'Combined_Key', 'Population')) %>%
  pivot_longer(cols = -c('Admin2','Province_State', 'Lat', 'Long_'),
               names_to = 'Date',
               values_to = 'deaths') %>%
  mutate(Date = mdy(Date))

head(us_deaths)

```

```{r NA filtering}

# View summaries to better understand the data
summary(us_cases)
summary(us_deaths)

# Filter out where cases/deaths are less than zero
us_cases %>%
  filter(cases < 0)

us_deaths %>% 
  filter(deaths < 0)

# Filter around these dates to see where it's happening
start_date <- '2022-05-03'
end_date <- '2022-05-10'

us_cases %>% 
  filter(Province_State == "South Carolina" & Date >= start_date & Date <= end_date)

us_deaths %>% 
  filter(Province_State == "South Carolina" & Date >= start_date & Date <= end_date)

# Filter out out where admin is 'Unassigned' because all values either 0 or neg
us_cases <- us_cases %>%
  filter(Admin2 != 'Unassigned')

us_deaths <- us_deaths %>% 
  filter(Admin2 != 'Unassigned')
```
```{r summaries}
# Check that worked as needed
summary(us_cases)
summary(us_deaths)
```



```{r us data join}
# Join all the data together into one table
us_data <- us_cases %>%
  full_join(us_deaths)

# Check the column names
colnames(us_data)

# See if there are any NA values
na_counts <- colSums(is.na(us_data))
print(na_counts)

# Sanity check that data all lines up
summary(us_data)

```
# Data Analysis and Visualizations

```{r total calculations}
# Calculate max date, total cases, and total deaths

max_date <- max(us_data$Date)
max_date

us_total_cases <- us_data %>%
  filter(Date == max_date) %>%
  summarize(total_cases =sum(cases))

us_total_deaths <- us_data %>%
  filter(Date == max_date) %>%
  summarize(total_deaths =sum(deaths))

print(us_total_cases)
print(us_total_deaths)
```


```{r cases per day}
# Calculate the total number of cases for each day
totals_summary <- us_data %>% 
  select(Date, deaths, cases) %>% 
  filter(Date >= '2020-01-22', Date <= max_date) %>% 
  group_by(Date) %>% 
  summarize(total_deaths = sum(deaths), total_cases = sum(cases))

head(totals_summary)
```

```{r cases/deaths over time graphs, warning = FALSE}
# Plot out the graph for the total cases and total deaths by date
ggplot(totals_summary, aes(x = Date, y = total_cases)) +
  geom_line(color = "blue", size = 1) +
  labs(title = "COVID-19 Cases Over Time",
       x = 'Date',
       y = 'Cases') +
  theme_minimal()

ggplot(totals_summary, aes(x = Date, y = total_deaths)) +
  geom_line(color = "red", size = 1) +
  labs(title = "COVID-19 Deaths Over Time",
       x = 'Date',
       y = 'Deaths') +
  theme_minimal()
```



In both of these charts, we can see the steady increase of cases and deaths over
time. This is expected because of the contagious aspect of the virus. Side by
side the charts can be slightly misleading because of the similar trends between
the two. However, without closer inspection of the y-axes, it could appear that
the number of cases and deaths are comparable to one another when in reality,
the number of cases drastically outweighs the number of deaths.The charts can be
easily interpreted, showing that there is a positive trend upwards with both
deaths and cases, graphically indicating a correlation between the cases and
deaths over time.


```{r total cases/deaths per state}

# Create a new data frame with the total cases for each state
totals_per_state <- us_data %>% 
  select(Province_State, deaths, cases) %>% 
  group_by(Province_State) %>% 
  summarize(total_deaths = max(deaths), total_cases = max(cases))

# Arrange in descending order to see the ones with the greatest total deaths
totals_per_state %>% 
  arrange(desc(total_deaths))

```

```{r totals per state charts}

# Make the Province_State column to a factor column
totals_per_state$Province_State <- factor(totals_per_state$Province_State, levels = totals_per_state$Province_State[order(totals_per_state$total_deaths)])

# Plot the bar chart with the descending order of the number of deaths and cases
# and filter just to the top 10
ggplot(head(totals_per_state %>% 
              arrange(desc(total_deaths)), 10), aes(x = Province_State, y = total_deaths, fill = total_deaths)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "pink", high = "red") + 
  coord_flip() +
  labs(title = "Total Deaths per State", x = "State", y = "Number of Deaths") +
  theme_minimal()

totals_per_state$Province_State <- factor(totals_per_state$Province_State, levels = totals_per_state$Province_State[order(totals_per_state$total_cases)])

# Plot the second bar chart in the same way
ggplot(head(totals_per_state %>% 
              arrange(desc(total_cases)), 10), aes(x = Province_State, y = total_cases, fill = total_cases)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "pink", high = "red") +  # You can remove this line
  coord_flip() +
  labs(title = "Total Cases per State", x = "State", y = "Number of Cases") +
  theme_minimal()


    
```



In these two graphs we can see which of the states were impacted the most in 
terms of both cases and deaths. California is significantly the most impacted in
both instances with states like Arizona and Florida and others are very highly
affected. It is possible that these states have a higher population count
which would help to understand why there are so many more cases and deaths.
By understanding which states can be significantly impacted by a rapidly spread
virus, this can help future public health officials to know which states may be 
the most at risk in the case of a future outbreak.


```{r population per state}
# Adding in the population element from the us_states_pop data
totals_per_state_and_pop <- totals_per_state %>% 
  full_join(us_states_pop) %>% 
  select('Province_State', 'total_deaths', 'total_cases', 'Population', 'Lat', 'Long_')

# View the data
head(totals_per_state_and_pop)
```


```{r cases per 100k}
# Calculate the number of cases for 100,000 people and create a new column
totals_per_state_and_pop <- totals_per_state_and_pop %>%
  mutate(cases_per_100k = round((total_cases / Population) * 100000),
         deaths_per_100k = round((total_deaths / Population) * 100000))

# Arrange in descending order to see the most affected states
head(totals_per_state_and_pop %>% 
  arrange(desc(deaths_per_100k
              )))
```


```{r us map, warning = FALSE}

# Read the GeoJSON file containing US states data into a SpatialPolygonsDataFrame
states <- geojsonio::geojson_read("https://rstudio.github.io/leaflet/json/us-states.geojson", what = "sp")

# Check the class of the 'states' object and inspect its names
class(states)
names(states)

# Create a Leaflet map object centered at latitude 37.8 and longitude -96
m <- leaflet(states) %>%
  setView(-96, 37.8, 4) %>%
  # Add MapBox tiles as the base layer
  addProviderTiles("MapBox", options = providerTileOptions(
    id = "mapbox.light",
    accessToken = Sys.getenv('MAPBOX_ACCESS_TOKEN')))



# Define the number of bins for the color palette
num_bins <- 10
# Generate breaks for the color bins based on 'cases_per_100k' data
breaks <- seq(min(totals_per_state_and_pop$cases_per_100k, na.rm = TRUE),
              max(totals_per_state_and_pop$cases_per_100k, na.rm = TRUE),
              length.out = num_bins + 1)

# Create a color palette with predefined bins
pal <- colorBin(palette = "YlOrRd", domain = totals_per_state_and_pop$cases_per_100k, bins = breaks)

# Add polygons representing states with color based on 'cases_per_100k'
m <- m %>%
  addPolygons(data = states,
              fillColor = ~pal(totals_per_state_and_pop$cases_per_100k),
              color = "black",
              weight = 2,
              opacity = 1,
              dashArray = "3",
              fillOpacity = 0.7,
              popup = ~paste("State: ", totals_per_state_and_pop$Province_State, "<br>",
                             "Cases per 100k: ", totals_per_state_and_pop$cases_per_100k, "<br>",
                             "Population: ", totals_per_state_and_pop$Population),
              # Options for highlighting a state on mouseover
              highlightOptions = highlightOptions(
                weight = 5,
                color = "#666",
                dashArray = "",
                fillOpacity = 0.7,
                bringToFront = TRUE))

# Render the map
m
```




This map demonstrates which states were most impacted with their populations 
being taken into account. Each color represents the percent of cases per 100k
people. This allows for us to understand that population does in fact play a 
large role in determining that the number of cases doesn't necessarily show the
overall impact on the state as it relates to its population.




```{r wisconsin data table}
# Filter the lookup_table to get population data for Wisconsin, 
# excluding rows with missing Admin2 (county) and Population values
wisconsin_pop_data <- lookup_table %>% 
  filter(Province_State == 'Wisconsin',
         !is.na(Admin2) & !is.na(Population))

# Print the filtered population data for Wisconsin
wisconsin_pop_data

# Calculate totals per county for deaths and cases in Wisconsin
totals_per_county <- us_data %>% 
  select(Admin2, Province_State, deaths, cases) %>%
  filter(Province_State == 'Wisconsin') %>% 
  group_by(Admin2, Province_State) %>% 
  summarize(total_deaths = max(deaths), total_cases = max(cases))

head(totals_per_county)
```

```{r population join, warning = FALSE}
# Perform a full join between totals_per_county and Wisconsin_pop_data
totals_per_county_and_pop <- totals_per_county %>% 
  full_join(wisconsin_pop_data) %>% 
  select('Admin2','Province_State', 'total_deaths', 'total_cases', 'Population', 'Lat', 'Long_')

# Filter out rows where Population is NA
totals_per_county_and_pop <- totals_per_county_and_pop[!is.na(totals_per_county_and_pop$Population), ]

head(totals_per_county_and_pop)
```

```{r totals per city}
# Add two new columns to totals_per_county_and_pop dataframe:
# cases_percent: percentage of total cases relative to population
# deaths_percent: percentage of total deaths relative to population
totals_per_county_and_pop <- totals_per_county_and_pop %>%
  mutate(cases_percent = (total_cases / Population) * 100,
         deaths_percent = (total_deaths / Population) * 100)


totals_per_county_and_pop %>%
  arrange(desc(cases_percent
              ))
```



```{r wisconsin map}
# Define breakpoints for creating bins based on cases_percent
breakpoints <- c(0,25,31,34,40,100)

# Create bins based on cases_percent with specified breakpoints
bins <- cut(totals_per_county_and_pop$cases_percent,
            breaks = breakpoints,
            include.lowest = TRUE)
# Define a color palette using the 'OrRd' Brewer palette
color_palette <- brewer.pal(length(breakpoints) - 1, "OrRd")

# Assign colors to data points based on their corresponding bins
colors <- cut(totals_per_county_and_pop$cases_percent,
              breaks = breakpoints,
              labels = color_palette)

# Create a Leaflet map object centered at longitude -89.5 and latitude 44.5
m <- leaflet() %>%
  setView(lng = -89.5, lat = 44.5, zoom = 7) %>% 
  addTiles()


# Add circle markers to the map representing each county
m <- m %>% 
  addCircleMarkers(data = totals_per_county_and_pop,
                   lng = ~Long_,
                   lat = ~Lat,
                   radius = ~sqrt(Population) / 25,  # Adjust radius based on population
                   color = ~colors,
                   fillOpacity = 0.5,
                   popup = ~paste("City: ", Admin2, "<br>", 
                                  "Percent of Population affected: ", cases_percent))

# Add a legend to the map displaying the color scale and corresponding labels
m <- m %>%
  addLegend("bottomright", title = "Cases Percent", colors = color_palette, 
            labels = c('0-25','26-31','32-34','35-49', '50+'))
# Display the map
m
```


This final map shows a spread of cases throughout the state of Wisconsin. This 
gives us a closer look at the highly concentrated locations. The size of the
indicates the population size and the color represents the percent of cases
per 100,000 people. According to the map, the sizes of the circles appear to
have a correlation with the color of the circle, however this is not exclusively
the case as there is a significant outlier such as Menominee with a small
population but was 52% for the population affected. Madison also has one of the
largest populations but is not the most highly impacted.



# Results

The findings suggest that the size of the population appears to have an affect
on the overall number of cases but this doesn't mean that the majority of the
population was even affected. We can see the steady increase in both cases and 
deaths, with a clear and somewhat obvious correlation between the number of 
cases and how that would lead to more deaths by the virus.

Some limitations of this data set and analysis could be that the same individuals
could have multiple cases of the virus since the data provided is over several
years. This could change the percent of the population is affected. We also are
not able to see the spread of the virus as it relates to the age or health
demographics of the areas to see if that would be a factor in the spread in a 
city/state. One other limitation of this data set is understanding that it is 
possible that not everyone who even had COVID, reported it. Access to COVID 
tests can also impact how many people were able to even report that they had 
the virus.


# Conclusion

In conclusion, our analysis has provided valuable insights into the dynamics of 
COVID-19 transmission within the United States. We found that the rate of
infection varies significantly across different states and regions,
with densely populated areas experiencing higher case counts compared to other
areas.

Our analysis highlights the need for equitable distribution of resources and
support to communities disproportionately affected by the pandemic.
Looking ahead, continued monitoring and analysis of COVID-19 trends are
necessary to adapt public health policies and interventions effectively.
By leveraging data-driven insights and collaborative efforts, we can work
towards controlling the spread of the virus and safeguarding the health
and well-being of communities nationwide.


